\textbf{Handbook on ``Entropy, Complexity, and Spatial Dynamics: The Rebirth of Theory?''}\medskip\\ Chapter by Juste Raimbault\textsuperscript{1} and Denise Pumain\textsuperscript{2}\medskip\\ \textbf{Spatial dynamics of complex urban systems within an evolutionary theory frame}

### 2021-01-29 Docs/Papers/HandbookEntropy/chapter_v6_20201214.tex
handbook_entropy

\title{Spatial dynamics of complex urban systems within an evolutionary theory frame}

@@

\author{
Juste Raimbault\textsuperscript{1,2,3*},
Denise Pumain\textsuperscript{3}\bigskip\\
\textsuperscript{1} Centre for Advanced Spatial Analysis, UCL, London, United Kingdom
\\
\textsuperscript{2} UPS CNRS 3611 ISC-PIF, Paris, France
\\
\textsuperscript{3} UMR CNRS 8504 G{\'e}ographie-cit{\'e}s, Paris, France\bigskip\\
* \texttt{juste.raimbault@polytechnique.edu}
}

@@

	\textbf{Keywords:} spatial dynamics; complex systems; system of cities; evolutionary theory; urban growth; simulation
\end{abstract}

@@

The context of this chapter, namely ``Entropy, Complexity, and Spatial Dynamics: The Rebirth of Theory'' is challenging. The problem is twofold: first, it means demonstrating how the insights provided by concepts and models associated with the notion of complexity have contributed to theories on spatial dynamics; second, it questions whether this is a rebirth of the theoretical approach in this field, which would therefore presume its relative abandon before.

This chapter provides answers to these two questions about a particular dimension of spatial dynamics, that of cities and systems of cities. In response to the second question, and considering the progress of research on urban complexity over the last forty years or so, we identify an enrichment and consolidation of existing theories by the principles of complex systems rather than a true renaissance of urban theory. Indeed, contrary to what some scholars claim, the theories developed around the spatial dynamics of urbanization processes and the evolution of settlement systems are far from being obsolete \citep{brenner2014urban}. Neither are they limited in their fundamentals and application to only one part of the world \citep{robinson2016comparative}. \cite{scott2015nature} and \cite{wu2020emerging} rightly criticized both ``theories''. We may add that many scholars did not abandon the former urban theories but on contrary tested, revised and completed them during the last four decades \citep{pumain1997pour,pumain1998urban,pumain2003approche,pumain2020theories,batty2013new}. The major theoretical challenge of that period was three folds: first, to shift from idiographic postures to nomothetic ones; second, from static views toward dynamic and evolutionary; and third, to really transfer concepts and models from natural sciences to build a socially relevant knowledge. Elements of knowledge from ancient theories can consolidate when situated in their geo-historical contexts, when tests of these theoretical propositions are carried out on subsets of empirical data that are well identified in time and space, and when the means are provided to articulate the observations made at multiple levels of geographic scales.

Starting from such a spiral theoretical conception of the cumulativeness of knowledge \citep{pumain2005cumulativite}, we try here to demonstrate how it is possible to construct an evolutionary theory of cities and systems of cities. This theory is abductive in the sense that it is constructed by frequent to-and-fro movements between empirical observations, logical propositions and mathematical and computer models. This theory is not entirely new as it integrates elements of knowledge already well identified by specialists in urban issues, such as geographers, economists, historians and archaeologists, sociologists, town planners and architects. Its originality stems from progress in international comparisons made through using new harmonized databases, methods for validating the results of computer simulation models, and putting these results into the perspective of complex systems theories.

@@

Two repeated observations prompted the development of a theory of the evolution of systems of cities. The first, empirical, is the long persistence of urban networks in large integrated territories, which retain, over time, the same spatial configurations of the relative size of cities. This is visible for example on maps representing the population of cities in proportional circles more than a century apart, in Europe, in India, or even in the United States since 1950. The second results of applying to the evolution of these urban population distribution a simple stochastic statistical model formalized by the French statistician Gibrat since 1931. This model states that the growth of urban populations is proportional to the initial size of cities during each short time interval, with fluctuations such as growth rates (amount reported to the population) are statistically independent from the size of the cities and from one time interval to the next. The interest of this model of spatially distributed urban growth in integrated systems of cities is to predict that the result of such a growth process is always a lognormal distribution. Thus, the model provides a first statistical explanation to the ``mystery'' of the Zipf's law still recently mentioned by the economist Paul \cite{krugman1996confronting}. A huge literature is dedicated to Zipf’s law and Pareto distributions that are observed in many natural and social systems, to the extent that these ``universal'' laws, together with fractal spatial patterns, have been sometimes considered as the ``signature'' of complexity in different systems. The underlying regularities in the urban spatial dynamics that these models capture have attracted many scientists from various disciplines to propose different theories of urban complexity.

@@

This did not prevent too many experimented and incidentally prominent scholars to capture “urban data” without any care about their quality, selection, claiming and striking results after crunching them with sophisticated statistical or mathematical models. Such an attitude is detrimental, for instance regarding the endless controversies about the models of city sizes distribution, for building a sound science that accumulates knowledge with reproducible means \citep{pumain2012theorie}. The resulting cacophony is sometimes disentangled in courageous articles, such as for instance \citep{cottineau2017metazipf,cottineau2020metametazipf} in a meta-review of hundreds of references, but certainly represents a waste of time and efforts that could be avoided. A recent paper seemed promising in announcing the observation of worldwide trends in urbanization from 1950 to 2030 using some 1857 cities larger than 300 000 inhabitants from 155 countries \citep{egidi2020long}. The paper offers surprising results such as a systematic ``inverse relationship'' between urban growth and city size in all regions ``up to the late 1990s''. At first sight, this could result from mixing countries with opposite urban growth trends, but the authors observe it as well in homogenous sub regions. However, after questioning the authors who kindly accepted to precise the origin of their data, it seems that the selected cities were delineated over time in a way not measuring the spatial expansion of the urban area. This could explain why the results seem in contradiction with those obtained from large parts of the world during the same period with harmonized urban databases \citep{cura2017old}. It also reminds about the necessity of starting with a sound conceptual definition of cities for measuring their attributes and controlling the possible effects of the data selection on the results. The urban definition may vary but should be in accordance with concepts and theory, i.e. with the research question and the selection of attributes that are measured for answering them \citep{rozenblat2020extending}.

@@

Although the recent development of networks and the accelerating circulation of information seem to blur the three-level conceptual definition of these geographical objects, we believe that this representation remains useful and relevant for guiding urban research and applications. An interesting contribution by C. \cite{roth2006reconstruction} discusses the emergentist view compared to the reductionist and suggests that different modes of access and not only bottom- up interactions could be part of the generative processes of the features constituting one level – which is the view point from statistical mechanics in envisaging self-organizing systems. This rejoins our theory for which urban interactions shaping cities and systems of cities are mostly but not exclusively conceived as emanating from one low level toward the higher level, in societal processes that take a long time, often longer than a human life. Over decades and sometimes centuries, most frequent interactions among urban stakeholders at micro level shape the emerging properties of a city at meso-geographical level, and inter-urban interactions generate the hierarchical organization and the functional differentiation of cities within the system of cities at macro-level. However, for instance, a location decision taken by an individual firm (at micro-level) can operate a definite change in the relative specialization of a particular city within the system of cities (at macro level), as well as many policy regulations decided at macro-level may directly change the situation of an individual firm or citizen at micro-level inside a city.

@@

While trying to identify major processes guiding the evolution of cities within system of cities, we stressed the role of information and innovation \citep{lane2009complexity}. Innovation when defined as a socially accepted invention may represent a form of social organization or a belief or a practice as well as a technological device or a new service. We shall develop further in the models presented below how innovation and its more or less always hierarchical diffusion among cities within systems of cities appears as a major dynamic process for explaining the expansion of urbanization and the persistency of cities’ relative situation in urban systems.

@@

@@

@@

>>>>

\title{\textbf{Handbook on ``Entropy, Complexity, and Spatial Dynamics: The Rebirth of Theory?''}\medskip\\
Chapter by Juste Raimbault\textsuperscript{1} and Denise Pumain\textsuperscript{2}\medskip\\
\textbf{Spatial dynamics of complex urban systems within an evolutionary theory frame}}

@@

%\author{
%Juste Raimbault\textsuperscript{1,2,3*},
%Denise Pumain\textsuperscript{3}\bigskip\\
%\textsuperscript{1} Centre for Advanced Spatial Analysis, UCL, London, United Kingdom
%\\
%\textsuperscript{2} UPS CNRS 3611 ISC-PIF, Paris, France
%\\
%\textsuperscript{3} UMR CNRS 8504 G{\'e}ographie-cit{\'e}s, Paris, France\bigskip\\
%* \texttt{juste.raimbault@polytechnique.edu}
%}

\author{\textsuperscript{1}Research Fellow, Center for Advanced Spatial Analysis,\\
University College London. ORCID: 0000-0003-0768-9480\\
\textsuperscript{2} Emeritus Professor, Universit{\'e} Paris I Panth{\'e}on-Sorbonne,\\
UMR G{\'e}ographie-cit{\'e}s. ORCID : 0000-0002-0730-4563}

@@

	\textbf{Keywords: spatial dynamics; complex systems; system of cities; evolutionary theory; urban growth; simulation}
\end{abstract}

@@

The title chosen for this book: ``Entropy, Complexity, and Spatial Dynamics: The Rebirth of Theory'' is challenging. The problem is twofold: first, it means demonstrating how the insights provided by concepts and models associated with the notion of complexity have contributed to theories on spatial dynamics; second, it questions whether this is a rebirth of the theoretical approach in this field, which would therefore presume its relative abandon before.

This chapter provides answers to these two questions about a particular dimension of spatial dynamics, that of cities and systems of cities. In response to the second question, and considering the progress of research on urban complexity over the last forty years or so, we identify an enrichment and consolidation of existing theories by the principles of complex systems rather than a true renaissance of urban theory. Indeed, contrary to what some scholars claim, the theories developed around the spatial dynamics of urbanization processes and the evolution of settlement systems are far from being obsolete \citep{brenner2014urban}. Neither are they limited in their fundamentals and application to only one part of the world \citep{robinson2016comparative}. \cite{scott2015nature} and \cite{wu2020emerging} rightly criticized both ``theories''. We may add that many scholars did not abandon the former urban theories but on contrary tested, revised and completed them during the last four decades \citep{pumain1997pour,pumain1998urban,pumain2003approche,pumain2020theories,batty2013new}. The major theoretical challenge of that period was three fold: first, to shift from idiographic postures to nomothetic ones; second, from static views toward dynamic and evolutionary; and third, to really transfer concepts and models from natural sciences to build a socially relevant knowledge. Elements of knowledge from ancient theories can consolidate when situated in their geo-historical contexts, when tests of these theoretical propositions are carried out on subsets of empirical data that are well identified in time and space, and when the means are provided to articulate the observations made at multiple levels of geographic scales.

Starting from such a spiral theoretical conception of the cumulativeness of knowledge \citep{pumain2005cumulativite}, we try here to demonstrate how it is possible to construct an evolutionary theory of cities and systems of cities. This theory is abductive in the sense that it is constructed by frequent to-and-fro movements between empirical observations, logical propositions and mathematical and computer models. This theory is not entirely new as it integrates elements of knowledge already well identified by specialists in urban issues, such as geographers, economists, historians, archaeologists, sociologists, town planners and architects. Its originality stems from progress in international comparisons made through using new harmonized databases, methods for validating the results of computer simulation models, and putting these results into the perspective of complex systems theories.

@@

Two repeated observations prompted the development of a theory of the evolution of systems of cities. The first, empirical, is the long persistence of urban networks in large integrated territories, which retain, over time, the same spatial configurations of the relative size of cities. This is visible for example on maps representing the population of cities in proportional circles more than a century apart, in Europe, in India, or even in the United States since 1950. The second results of applying to the evolution of these urban population distribution a simple stochastic statistical model formalized by the French statistician Gibrat since 1931. This model states that the growth of urban populations is proportional to the initial size of cities during each short time interval, with fluctuations such as growth rates (amount reported to the population) are statistically independent from the size of the cities and from one time interval to the next. The interest of this model of spatially distributed urban growth in integrated systems of cities is to predict that the result of such a growth process is always a lognormal distribution. Thus, the model provides a first statistical explanation to the ``mystery'' of the Zipf's law \citep{zipf1941national} still recently mentioned by the economist Paul \cite{krugman1996confronting}. A huge literature is dedicated to Zipf’s law and Pareto distributions that are observed in many natural and social systems, to the extent that these ``universal'' laws, together with fractal spatial patterns, have been sometimes considered as the ``signature'' of complexity in different systems. The underlying regularities in the urban spatial dynamics that these models capture have attracted many scientists from various disciplines to propose different theories of urban complexity.

@@

This did not prevent too many experimented and incidentally prominent scholars to capture “urban data” without any care about their quality, selection, claiming and striking results after crunching them with sophisticated statistical or mathematical models. Such an attitude is detrimental, for instance regarding the endless controversies about the models of city sizes distribution, for building a sound science that accumulates knowledge with reproducible means \citep{pumain2012theorie}. The resulting cacophony is sometimes disentangled in courageous articles, such as for instance \citep{cottineau2017metazipf,cottineau2020metametazipf} in a meta-review of hundreds of references, but certainly represents a waste of time and efforts that could be avoided. A recent paper seemed promising in announcing the observation of worldwide trends in urbanization from 1950 to 2030 using some 1857 cities larger than 300,000 inhabitants from 155 countries \citep{egidi2020long}. The paper offers surprising results such as a systematic ``inverse relationship'' between urban growth and city size in all regions ``up to the late 1990s''. At first sight, this could result from mixing countries with opposite urban growth trends, but the authors observe it as well in homogenous sub regions. However, after questioning the authors who kindly accepted to precise the origin of their data, it seems that the selected cities were delineated over time in a way not measuring the spatial expansion of the urban area. This could explain why the results seem in contradiction with those obtained from large parts of the world during the same period with harmonized urban databases \citep{cura2017old}. It also reminds about the necessity of starting with a sound conceptual definition of cities for measuring their attributes and controlling the possible effects of the data selection on the results. The urban definition may vary but should be in accordance with concepts and theory, i.e. with the research question and the selection of attributes that are measured for answering them \citep{rozenblat2020extending}.

@@

Although the recent development of networks and the accelerating circulation of information seem to blur the three-level conceptual definition of these geographical objects, we believe that this representation remains useful and relevant for guiding urban research and applications. An interesting contribution by Camille \cite{roth2006reconstruction} discusses the emergentist view compared to the reductionist and suggests that different modes of access and not only bottom- up interactions could be part of the generative processes of the features constituting one level – which is the view point from statistical mechanics in envisaging self-organizing systems. This rejoins our theory for which urban interactions shaping cities and systems of cities are mostly but not exclusively conceived as emanating from one low level toward the higher level, in societal processes that take a long time, often longer than a human life. Over decades and sometimes centuries, most frequent interactions among urban stakeholders at micro level shape the emerging properties of a city at meso-geographical level, and inter-urban interactions generate the hierarchical organization and the functional differentiation of cities within the system of cities at macro-level. However, for instance, a location decision taken by an individual firm (at micro-level) can operate a definite change in the relative specialization of a particular city within the system of cities (at macro level), as well as many policy regulations decided at macro-level may directly change the situation of an individual firm or citizen at micro-level inside a city.

@@

While trying to identify major processes guiding the evolution of cities within system of cities, we stressed the role of information and innovation \citep{lane2009complexity}. Innovation when defined as a socially accepted invention may represent a form of social organization or a belief or a practice as well as a technological device or a new service \citep{broitman2020endogenous}. We shall develop further in the models presented below how innovation and its more or less always hierarchical diffusion among cities within systems of cities appears as a major dynamic process for explaining the expansion of urbanization and the persistency of cities’ relative situation in urban systems.

@@

%\bibliographystyle{apalike}
%\bibliography{biblio}

@@

\bibitem[Broitman and Czamanski, 2020]{broitman2020endogenous}
Broitman, D. and Czamanski, D. (2020).
\newblock Endogenous growth in a spatial economy: The impact of globalization
  on innovations and convergence.
\newblock {\em International Regional Science Review}, page 0160017620946095.

@@

\bibitem[Zipf, 1941]{zipf1941national}
Zipf, G. (1941).
\newblock {\em National unity and disunity}.
\newblock Bloomington Principia Press (Indiana).
